Generic Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-03-29 latex/intro.tex
Fixin

Designing experiments is usually done by optimizing some \emph{design
criterion}. This is true both for frequentists
\cite{silvey2013,Ucinski05} as well as for Bayesians
\cite{Chaloner1995}. See \cite{Chaloner1995} for an investigation of
the analogy between the two approaches. Although there is a plethora
of design criteria, we focus on just one of these, commonly referred
to as \emph{D-optimal design}. It has a simple and appealing
motivation in the Bayesian context as explained in
\cite{Chaloner1995}: Consider a linear model in finite dimensions,
with Gaussian prior and noise. Under this model, maximizing the
expected information gain (KL divergence \cite{CoverThomas91}) between
posterior and prior amounts to minimizing the determinant of the
posterior covariance matrix. In a frequentist setting, a D-optimal
design minimizes the volume of the uncertainty ellipsoid \cite[page
  16]{Ucinski05}, but this is done for the Fisher information matrix
and not the posterior covariance. However, \cite{Chaloner1995} shows
that the latter is just a regularized version of the former.

The previous discussion is classical for experimental design when
inference takes place over a finite (not too large) number of
parameters. The subject of optimal experimental design for function
inference in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, it can be shown that a D-optimal design arises naturally for
linear models when one wishes to maximize the KL divergence between
posterior and prior. This amounts to minimizing the determinant of the
posterior covariance operator (understood as a product of its
eigenvalues). Some difficulties arise in the process, but remedies can
be found as shown in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D criterion, the optimization process results in
observations that are very similar. For example, if an observation is
thought of as measuring some function value at $\x \in \Omega
\subseteq \R^d, d=1,2,3$ (with added error) then the optimization
procedure sometimes places sensors in very close proximity to each
other (as can be seen in figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{fedorov1996} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design.
In such designs, the user enforces observation locations to be far
from each other. One way to do this is by introducing correlated
errors which, philosophically, accounts for both observation error and
model error. Another method considered is imposing distance
constraints between observations. A somewhat different approach is
suggested in \cite[page 49]{fedorov2012}, where close by design
observations are merged --- a procedure which obviously does not avoid
clusterization. The same problem arises in time-series analysis for
pharmacokinetic experiments. The authors of \cite{hooker2009} suggest
modeling auto-correlation time in the noise model, which is equivalent
to the correlated errors mentioned above.

@@

Similarly, in the inverse problems community, work is mostly
computational and less theoretic. Model errors were considered in
\cite{attia2020, koval2020}. The former study is focused on inferring a
Quantity of Interest (QoI). The focus of the latter is reducing
forward solves, using randomized linear algebra. Both studies present
numerical techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. However, neither of these studies mention any effect model
errors can have on sensor clusterization. The current study is mostly
theoretical and aims to fill the gap of understanding sensor
clusterization.

\subsection{Contribution}
We propose and thoroughly study a relaxed and analytically tractable
model for understanding D-optimal designs. Under this model, D-optimal
designs are solutions of a constrained optimization problem,
formulated using Lagrange multipliers (section \ref{section:D and
  grad}). This allows us to rigorously show how model error mitigates
clusterization (section \ref{section:non vanishing}). We show that
designs that exhibit clusterization are just as optimal as
clusterization free designs.

A beautiful mathematical structure arises in D-optimal designs when no
model error is present. The Lagrange multipliers problems is in fact a
nonlinear eigenvalue problem for the observations and prior
covariance. The operator for which eigenvectors and eigenvalues are
sought is a sum of two operators. The first is the prior
covariance. The second is an outer product of the observations (see
section \ref{section:vanishing} for details and exact statement). This
structure helps us shed light on D-optimal designs. We characterize
these designs in Theorem \ref{thm:char}. One insight we prove is that
a D-optimal design reduces uncertainties where they are highest
first. Other interestig phenomena arise, but they require setting
notation, and are discussed later.

In the process, we generalize several lemmas from linear algebra to
infinite-dimensional settings. We prove a Matrix Determinant Lemma in
\ref{lemma:MDL}. We generalize a lemma due to Lax \cite{Lax07} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. We also prove
lemmas in linear algebra. One constructs a decomposition $M = AA^t$
where $A$ has unit norm columns \ref{lemma:free}. Another shows
simultaneous diagonizability of the prior and outer product in the
nonlinear eigenvalue problem referenced above \ref{lemma:sim diag}. We
also provide other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems.%%  Among these are tools
%% for calculating the increase in the design criterion per observation
%% --- Lemma \ref{lemma:design increase} and Corollary \ref{cor:zero mod err}.

@@

\subsection{Limitations}\label{subsec:limitations}
There are two main drawbacks of the study presented here. Our relaxed
model does not consider any specific set of allowed
observations. Rather, we take observations in the unit ball in some
Hilbert space. This allows considerably less restrictive observations
than any real-life problem does. The second drawback is that we do not
show rigorously that clusterization necessarily occurs in a
simultaneous design. We only show that it is as reasonable as no
clusterization.

@@

In section \ref{section:prelim} we present a more abstract and general
formulation of the inverse problem we consider. But, for the purpose
of illustration, we present clusterization via a toy model --- the 1D
heat equation in $[0,\pi]$ with a homogeneous Dirichlet boundary
condition.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1,
\dots,m$ and a final time $T > 0$. We assume centered Gaussian
observation error, so we observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$
with $\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid. We model
the initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. If It
is well known \cite{Tarantola05} that for linear problems, with
Gaussian prior and error, the posterior is also Gaussian with a
covariance that does not depend on the observed data. The posterior
covariance $\postcov$ is known to have a closed form formula, even in
infinite dimensions \cite{Stuart10}. We denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. The
posterior covariance is known and depends only on $\prcov, \fwd, \obs$
and $\sigma^2$ (see section \ref{section:prelim} and
\eqref{eq:postcov} specifically).

>>>>

Designing experiments is usually done by optimizing some \emph{design
criterion}. This is true both for frequentists
\cite{silvey2013,Ucinski05} as well as for Bayesians
\cite{Chaloner1995} (see \cite{Chaloner1995} for an investigation of
the analogy between the two approaches). Among the plethora of design
criteria, \emph{D-optimal design} have one of the simplest and most
appealing motivation in the Bayesian context \cite{Chaloner1995}: a
D-optimal design maximizes the expected information gain (KL
divergence \cite{CoverThomas91}) between posterior and prior. For a
linear model in finite dimensions, with Gaussian prior and noise, a
D-optimal design minimizes the determinant of the posterior covariance
matrix. In a frequentist setting, a D-optimal design minimizes the
volume of the uncertainty ellipsoid \cite[page 16]{Ucinski05}, but
this is done for the Fisher information matrix and not the posterior
covariance. However, \cite{Chaloner1995} show that the latter is just
a regularized version of the former.

The previous discussion is classical for experimental design when
inference takes place over a finite (not too large) number of
parameters. The subject of optimal experimental design on function
spaces in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design still maximizes the KL divergence between
posterior and prior. For a linear model with Gaussian prior and noise,
this amounts to minimizing the determinant of the posterior covariance
operator (understood as a product of its eigenvalues). Some
difficulties arise in the process, but remedies can be found as shown
in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D-optimality criterion, the optimization process results in
observations that are very similar. For example, if an observation is
thought of as measuring some function value at $\x \in \Omega
\subseteq \R^d, d=1,2,3$ (with added error) then the optimization
procedure sometimes places sensors in very close proximity to each
other (as can be seen in figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{fedorov1996} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design.
In such designs, the user enforces observation locations to be far
from each other. One way to do this is by introducing correlated
errors which account for both observation error and model
error. Another method considered is imposing distance constraints
between observations. A somewhat different approach is suggested in
\cite[page 49]{fedorov2012}, where close observations are merged --- a
procedure which obviously does not avoid clusterization. The same
problem arises in time-series analysis for pharmacokinetic
experiments. The authors of \cite{hooker2009} suggest modeling
auto-correlation time in the noise model, which is equivalent to the
correlated errors mentioned above.

@@

In the inverse problems community, work is mostly computational and
less theoretic. Model errors were considered in \cite{attia2020,
  koval2020}. The former study is focused on inferring a Quantity of
Interest (QoI). The focus of the latter is reducing forward solves,
using randomized linear algebra. Both studies present numerical
techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem by
taking a Laplace approximation for the posterior). Both find an
optimal design by first solving a continuous problem for sensor
weights. Said solution is then sparsified to give a binary
design. Both studies are successful in the task of Bayesian
inversion. However, neither of these studies mention any effect model
errors can have on sensor clusterization. This study is mostly
theoretical and aims to fill the gap of understanding sensor
clusterization.

\subsection{Contribution}
I propose and study a relaxed and analytically tractable model for
D-optimal designs. Under this model, D-optimal designs are solutions
of a constrained optimization problem, formulated using Lagrange
multipliers. This allows me to rigorously show how model error
mitigates clusterization (section \ref{section:non vanishing}). I show
that designs that exhibit clusterization are indeed optimal.

%% A beautiful mathematical structure arises in D-optimal designs

When no model error is present, the Lagrange multipliers problem boils
down to a nonlinear eigenvalue problem.
%% The operator for which eigenvectors and eigenvalues are sought is a
%% sum of two operators. The first is the prior covariance. The second
%% is an outer product of the observations (see section
%% \ref{section:vanishing} for details and exact statement).
This structure helps me characterize D-optimal designs in Theorem
\ref{thm:char}. The key idea is that a D-optimal design reduces
uncertainties where they are highest first.

@@

In the process, I generalize several lemmas from linear algebra to
infinite-dimensional settings. I prove a Matrix Determinant Lemma in
\ref{lemma:MDL}. I generalize a lemma due to Lax \cite{Lax07} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. I show how to
decompose a symmetric positive definite matrix $m$ as $M = AA^t$,
where $A$ has unit norm columns in Lemma \ref{lemma:free}.

\subsection{Limitations}\label{subsec:limitations}
Two main drawbacks of the study presented here are: First, the relaxed
model does not consider any specific set of allowed
observations. Rather, I take observations in the unit ball in some
Hilbert space. This allows considerably less restrictive observations
than any real-life problem does. The second drawback is that I do not
show rigorously that clusterization necessarily occurs, I only show
that it is as reasonable as no clusterization.

@@

In section \ref{section:prelim} I present a more abstract and general
formulation of the inverse problem I consider. But, for the purpose of
illustration, I present clusterization via a toy model --- the 1D heat
equation in $[0,\pi]$ with a homogeneous Dirichlet boundary condition.

@@

The goal is to infer the initial condition $u_0$. For that purpose, we
measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian observation
error, so we observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2)$ iid, $\sigma^2 > 0$. We model
the initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. It is
known \cite{Tarantola05} that for linear problems, with Gaussian prior
and error, the posterior is also Gaussian with a covariance that does
not depend on the observed data. The posterior covariance $\postcov$
has a closed form formula, even in infinite dimensions
\cite{Stuart10}. We denote by $\fwd$ the dynamics operator, so that
$u( \cdot,T) = \fwd u_0$, and the observation operator $\obs$ so that
$u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. The posterior covariance is
known and depends only on $\prcov, \fwd, \obs$ and $\sigma^2$ (see
section \ref{section:prelim} and \eqref{eq:postcov} specifically).

### 2023-03-29 latex/paper.tex
Fixin

This material is based upon work supported by the National Science
Foundation Graduate Research Fellowship Program under Grant No. (NSF
grant number). Any opinions, findings, and conclusions or
recommendations expressed in this material are those of the author(s)
and do not necessarily reflect the views of the National Science
Foundation. This work was also supported by The Raymond and Beverly
Sackler Post-Doctoral Scholarship.

>>>>

This research was supported in part by an appointment with the
National Science Foundation (NSF) Mathematical Sciences Graduate
Internship (MSGI) Program sponsored by the NSF Division of
Mathematical Sciences. This program is administered by the Oak Ridge
Institute for Science and Education (ORISE) through an interagency
agreement between the U.S. Department of Energy (DOE) and NSF. ORISE
is managed for DOE by ORAU. All opinions expressed in this paper are
the author's and do not necessarily reflect the policies and views of
NSF, ORAU/ORISE, or DOE.  This work was also supported by The Raymond
and Beverly Sackler Post-Doctoral Scholarship.
